Automatic word stress annotation of Russian running text (max 8 pages)

### 2015-01-07 rusaccent-nodalida-2015/rusaccent.tex
more
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58258 72bbbca6-d526-0410-a7d9-f06f51895060

We identify three different types of word stress ambiguity. \emph{Intraparadigmatic} ambiguity 
refers
to homographic wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

@@

The remaining two types of stress ambiguity occur between lexemes. \emph{Morphosyntactically
congruent homographs} are homographs that belong to separate lexemes, and whose morphosyntactic 
values are identical, as shown in Table~\ref{table:MScongruent}.

@@

\emph{Morphosyntactically incongruent homographs} are homographs that belong to separate
lexemes, and whose morphosyntactic values are different, as shown in 
\ref{ex:MSincongruent}.

@@

\rr{another relevant issue is whether the ambiguity can be resolved at all. Maybe this is 
something we should bring up in the discussion section. For example, when is stress marked in 
authentic by-native-for-native texts? A computer should not be expected to predict the stress on 
such words (although a computer could possibly identify words that should be marked for 
natives.)} It should be noted that not all morphosyntactic ambiguity leads to stress ambiguity. For example,
all inflecting modifiers in Russian have the same form for F-SG-GEN, F-SG-LOC, F-SG-DAT and 
F-SG-INS. We refer to this as \emph{stress-irrelevant} ambiguity, since all readings have
the same stress placement. Stress-irrelevant ambiguity can be intraparadigmatic and/or interlexical.

@@

Automatic stress placement in Russian is similar to Diacritic Restoration, a task
which has received increasing interest over the last 20 years. Generally speaking, diacritics
disambiguate otherwise homographic wordforms, so missing diacritics can
complicate many NLP tasks, such as text-to-speech and speech-to-text. For example, speakers 
of Czech may type emails and other communications
without standard diacritics. In order to generate speech from these texts, they must first be 
normalized by restoring diacritics. A slightly different
situation arises with Arabic and Hebrew, since standard orthography lacks vowel diacritics
altogether. For such languages, the `restoration' of vowel diacritics results in less
ambiguity than in standard orthography. For languages with inherently ambiguous orthography,
it may be more precise to refer to this as `diacritic enhancement', since it produces
text that is less ambiguous than the standard language. In this sense, Russian orthography
is similar to Arabic and Hebrew, since its vowel qualities are underspecified in
standard orthography.

Many studies of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. CITE Krivnova 
1998), but very few studies have targeted stress-marking itself, and they tend to
focus on place stress on unknown words, with almost no attention to the problems that
arise with known words. For example, 
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, CITE Hall \& Sproat (2013) trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms\rr{might it be worth computing the difference in stress
ambiguity between wordform lists and running text? i.e. what percentage of tokens have ambiguous
stress.}, so evaluations based on wordlists - as opposed to running
text - have inflated results. Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.

So far, the implicit target application the few studies related to automatic stress placement
in Russian has been text-to-speech and speech recognition technology. However, the target 
application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

@@

State-of-the-art part-of-speech and morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} We developed free
and open-source finite-state tools capable of analyzing and
generating stressed wordforms. A finite-state transducer\footnote{a two-level 
morphology, xfst/hfst, blah, blah, blah} generates all possible readings
of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our stress placement
system depends on the performance of the Constraint Grammar. Ideally, the Constraint Grammar would
successfully remove all but the correct reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

\begin{table*}[t]
  \centering
  \begin{tabular}{r|c|c}
    & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} \\
    & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} \\
    & \rus{костить}-V-IPFV-IMP\hskip 1em\rus{кост\'{и}} & \\
    \hline
    {\small {\tt bare}} & \rus{кости} & \rus{кости} \\
    {\small {\tt safe}} & \rus{кости} & \rus{к\'{о}сти} \\
    {\small {\tt guess}} & \rus{к\'{о}сти} or \rus{кост\'{и}} & \rus{к\'{о}сти}\\
    {\small {\tt guessFreq}} & \rus{к\'{о}сти} & \rus{к\'{о}сти}
  \end{tabular}
  \caption{Example output of each condition, given a particular set of readings}
  \label{tab:conditions}
\end{table*}

@@

For our baseline, we take the output of our morphological analyzer (without the Constraint 
Grammar) in combination with {\small {\tt bare}}, {\small {\tt safe}}, {\small {\tt guess}} and {\small {\tt guessFreq}}. The purpose of this study is to compare the

\subsection{Electronically-available resources}

Nodalida-2015 provides this description in \LaTeX2e ({\small {\tt
    nodalida2015.tex}}) and PDF format ({\small {\tt nodalida2015.pdf}}),
along with the \LaTeX2e style file used to format it ({\small {\tt
    nodalida2015.sty}}) and an ACL bibliography style ({\small {\tt
    acl.bst}}) in case you want to use the Bib\TeX\ reference management
software. These files are all available at
\url{http://stp.lingfil.uu.se/~bea/nodalida15/}. A Microsoft Word template file ({\small
  {\tt nodalida2015.docx}}) is also available at the same URL. We strongly
recommend the use of these style files, which have been appropriately
tailored for the Nodalida-2015 proceedings. If you have an option, we
recommend that you use the \LaTeX2e version. \textbf{If you will be
  using the Microsoft Word template, we suggest that you anonymize
  your source file so that the pdf produced does not retain your
  identity.}  This can be done by removing any personal information
from your source document properties.

>>>>

We identify three different types of word stress ambiguity. \emph{Intraparadigmatic}
stress ambiguity refers
to homographic wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

@@

The remaining two types of stress ambiguity occur between lexemes. \emph{Morphosyntactically
congruent} stress ambiguity occurs between homographs that belong to separate lexemes, 
and whose morphosyntactic values are identical, as shown in Table~\ref{table:MScongruent}.

@@

\emph{Morphosyntactically incongruent} stress ambiguity occurs between homographs
that belong to separate lexemes, and whose morphosyntactic values are different, as shown in 
\ref{ex:MSincongruent}.

@@

\rr{another relevant issue is whether the ambiguity can be resolved at all. Maybe this is 
something we should bring up in the discussion section. For example, when is stress marked in 
authentic by-native-for-native texts? A computer should not be expected to predict the stress on 
such words (although a computer could possibly identify words that should be marked for 
natives.)} It should be noted that not all morphosyntactic ambiguity leads to stress ambiguity. For example, all inflecting modifiers have the same form for F-SG-GEN, F-SG-LOC, F-SG-DAT and 
F-SG-INS. We refer to this as \emph{stress-irrelevant} ambiguity, since all readings have
the same stress placement. Stress-irrelevant ambiguity can be intraparadigmatic and/or interlexical.

@@

Automatic stress placement in Russian is similar to Diacritic Restoration, a task
which has received increasing interest over the last 20 years. Generally speaking, diacritics
disambiguate otherwise homographic wordforms, so missing diacritics can
complicate many NLP tasks, such as text-to-speech. For example, speakers 
of Czech may type emails and other communications
without standard diacritics. In order to generate speech from these texts, they must first be 
normalized by restoring diacritics. A slightly different
situation arises with Arabic and Hebrew, since standard orthography lacks vowel diacritics
altogether. For such languages, the `restoration' of vowel diacritics results in less
ambiguity than in standard orthography. For languages with inherently ambiguous orthography,
it may be more precise to refer to this as `diacritic enhancement', since it produces
text that is less ambiguous than the standard language. In this sense, Russian orthography
is similar to Arabic and Hebrew, since its vowel qualities are underspecified in
standard orthography.

Many studies of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. CITE Krivnova 
1998), but very few studies have targeted stress-marking itself, and they tend to
focus on placing stress on unknown words, with almost no attention to the problems that
arise with known words. For example, 
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, CITE Hall \& Sproat (2013) trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms\rr{might it be worth computing the difference in stress
ambiguity between wordform lists and running text? i.e. what percentage of tokens have ambiguous
stress.}, so evaluations based on wordlists - as opposed to running
text - have inflated results. Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.

So far, the implicit target application the few studies related to automatic stress placement
in Russian has been text-to-speech and automatic speech recognition. However, the target 
application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

@@

State-of-the-art morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} To our knowledge,
no open-source resources are available for analyzing and generating stressed wordforms.
Therefore, we developed free and open-source finite-state tools capable of analyzing and
generating stressed wordforms. A finite-state transducer\footnote{Using two-level 
morphology, xfst/hfst, blah, blah, blah} generates all possible readings
of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our stress placement
system depends on the performance of the Constraint Grammar. Ideally, the Constraint Grammar would
successfully remove all but the correct reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

\begin{table*}[t]
  \centering
  \begin{tabular}{r|c|c}
    & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} \\
    & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} \\
    & \rus{костить}-V-IPFV-IMP\hskip 1em\rus{кост\'{и}} & \\
    \hline
    {\small {\tt bare}} & \rus{кости} & \rus{кости} \\
    {\small {\tt safe}} & \rus{кости} & \rus{к\'{о}сти} \\
    {\small {\tt guess}} & \rus{к\'{о}сти} or \rus{кост\'{и}} & \rus{к\'{о}сти}\\
    {\small {\tt guessFreq}} & \rus{к\'{о}сти} & \rus{к\'{о}сти}
  \end{tabular}
  \caption{Example output of each stress placement approach, given a particular set of readings}
  \label{tab:conditions}
\end{table*}

@@

For our baseline, we take the output of our morphological analyzer (without the Constraint 
Grammar) in combination with the {\small {\tt bare}}, {\small {\tt safe}}, 
{\small {\tt guess}} and {\small {\tt guessFreq}} approaches.

\subsection{Stress Corpus}

Obtaining a suitable corpus for this study Russian texts with marked word stress are relatively rare, except in materials
for language learners, which are predominantly proprietary.
